Schema Evolution in Interactive Programming Systems

### 2024-08-22 paper.tex
Related Work section

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. While Schema Change has been extensively studied in the context of databases~\cite{erhard06}, we observe that analogous problems occur in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in Related Work~\ref{sec:related}. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems that exemplify a diverse range of typical schema change scenarios which interrupt feedback loops. We believe these problems have not been completely solved in theory or practice (see Related Work~\ref{sec:related}). While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. We propose these challenge problems to motivate and facilitate further research. We hope they serve as examples to ground such research and as a basis for evaluating results. We invite contributions of further challenge problems and look forward to discussing competing solutions.

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (Appendix~\ref{sec:related}), past work
does not resolve all problems in our second challenge.

@@

We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type) that
defines the AST structure of the domain-specific language.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as type, code and data co-evolution as in the case of the Elm
architecture in Section~\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.

@@

\newpage
\appendix

\section{Literature Review}
\label{sec:related}

@@

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in Section \ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.

>>>>

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. While Schema Change has been extensively studied in the context of databases~\cite{erhard06}, we observe that analogous problems occur in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems that exemplify a diverse range of typical schema change scenarios which interrupt feedback loops. We believe these problems have not been completely solved in theory or practice (see \S\ref{sec:related}  Related Work). While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. We propose these challenge problems to motivate and facilitate further research. We hope they serve as examples to ground such research and as a basis for evaluating results. We invite contributions of further challenge problems and look forward to discussing competing solutions.

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), past work
does not resolve all problems in our second challenge.

@@

We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type) that
defines the AST structure of the domain-specific language.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as type, code and data co-evolution as in the case of the Elm
architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.

@@

\section{Related Work}
\label{sec:related}

@@

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in \S\ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.
